Using species distribution models to direct native seed collection efforts

### 2024-09-12 scripts/Manuscript.Rmd
plot of variable importance

A Species Distribution Model (alternatively referred to as 'Environmental niche models' etc.) uses presence, and optionally absence, records of a species and correlates these records to environmental variables (e.g. climate, land cover, and soil), using various families of statistical models (e.g. GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface. 
While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@a2022species). 
In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g. they proliferate on uncommon soils and are outcompeted on more typical soil types. 
Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored. 
Testing the applicability of SDMs to modeling the range of common species is further valuable as much biodiversity planning incorporates common species which are oftentimes viewed as the fabric holding together ecosystems, and as they may assist with planning restoration plantings under climate change.

While the goal of an SDM is to develop a testable hypothesis of where a self sustaining population of a species *can* grow, they generally lack an explicit link to estimating *where* a population will actually grow. 
We argue that where populations will occur is driven largely by dispersal limitations, rather than simply whether habitat is suitable. 
A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis. 
Connectivity analysis, as generally implemented in land management, often requires a variety of computationally intensive algorithms stemming from a variety of disciplines (e.g. circuit theory).
However we posit that in the case of widespread species useful results can be obtained by simply detailing the occupancy of drainage catchments, and the number of catchments between these occupied sites, and drainages with hypothesized potential habitat.

@@

## Species Occurrence Records  
Species presence records were collected from herbaria, citizen science initiatives (e.g. iNaturalist), and standardized ecological monitoring programmes (e.g. VegBank, Forest Inventory and Analysis) using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio]. 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

@@

|  No. Occupied Patches   |   Rank   | 
| :---------------------: | :------: | 
|           >= 3          |    5     |  
|           <= 2          |    6     |

@@

how many records removed manually?
how many records flagged by LDA?

Which variables were the most important rank positions across all models? 
How were model evaluation statistics?

@@

>>>>

A Species Distribution Model (alternatively referred to as 'Environmental niche models' etc.) uses presence, and optionally absence, records of a species and correlates these records to environmental variables (e.g. climate, land cover, and soil), using various families of statistical models (e.g. GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface. 
While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous [@a2022species]. 
In the few published instances where SDMs have been tested, they have oftentimes been for rare species which oftentimes have rather obvious and consequential drivers of their distributions, e.g. they proliferate on uncommon soils and are outcompeted on more typical soil types [@johnson2023field]. 
Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored. 
Testing the applicability of SDMs to modeling the range of common species is further valuable as much biodiversity planning incorporates common species which are oftentimes viewed as the fabric holding together ecosystems, and as they may assist with planning restoration plantings under climate change.

While the goal of an SDM is to develop a testable hypothesis of where a self sustaining population of a species *can* grow, they generally lack an explicit link to estimating *where* a population will actually grow [@guisan2005predicting)]. 
We argue that where populations will occur is driven largely by dispersal limitations, rather than simply whether habitat is suitable [@franklin2010moving]. 
A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis. 
Connectivity analysis, as generally implemented in land management, often requires a variety of computationally intensive algorithms stemming from a variety of disciplines (e.g. circuit theory).
However we posit that in the case of widespread species useful results can be obtained by simply detailing the occupancy of drainage catchments, and the number of catchments between these occupied sites, and drainages with hypothesized potential habitat.

@@

## Species Occurrence Records  
Species presence records were collected from herbaria, citizen science initiatives (e.g. iNaturalist), and standardized ecological monitoring programmes (e.g. VegBank, Forest Inventory and Analysis) using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio]. 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 1.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

@@

|  No. Occupied Patches   |   Rank   | 
| :---------------------: | :------: | 
|           >= 3          |    8     |  
|           <= 2          |    9     |

@@

```{r}
source('functions.R')
library(tidyverse)
```

```{r How many records were removed in the cleaning process?}

thinned <- bind_rows(
  sf::st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE),
  sf::st_read('../data/raw/occurrence/thinned/thinned_occurrences3.shp', quiet = TRUE)
) |>
  sf::st_drop_geometry() |>
  group_by(taxon) |>
  summarize(totalRecords = n())

pts2remove <- read.delim('../data/raw/species_manual_cleaning.txt', header = FALSE)
colnames(pts2remove) <- 'Species'

pos <- str_extract_all(pts2remove$Species, "\\([^()]+\\)")
pos <- lapply(pos, function(x){substring(x, 2, nchar(x)-1)})
exact <- lapply(pos, function(x){suppressWarnings({as.numeric(sub(' ' , '', unlist(strsplit(x,","))))})})
ranges <- lapply(pos, colon_balloon)

to_remove <- Map(c, exact, ranges)  
to_remove <- lapply(lapply(to_remove, na.omit), sort)
names(to_remove) <- stringr::str_remove(pts2remove$Species, ' c[(].*$')
recs_removed <- lapply(to_remove, function(x){out <- length(na.omit(gsub('^0$', NA, x)))}) |>
  cbind() |>
  data.frame() |>
  rownames_to_column() |>
  setNames(c('taxon', 'recordsRemoved')) |>
  mutate(recordsRemoved = as.numeric(recordsRemoved)) |>
  inner_join(thinned, by = 'taxon') |>
  mutate(propRemoved = if_else(recordsRemoved > 0, recordsRemoved/totalRecords, 0))

rm(exact, ranges, pos, pts2remove, to_remove, thinned)
rm(list = ls()[ sapply( ls(), exists, mode="function")])
```

```{r How many records flagged by LDA?}

f <- file.path('../results/lda', list.files('../results/lda'))
df <- do.call(rbind, lapply(f, function(x)
  cbind(read.csv(x), name= gsub('[.]csv$', '', basename(x))))) |>
  mutate(
    removedFlagged = if_else(Removed >= Max_Recs_Removal, 1, 0),
    propFlagged = Flagged/Abs_in
    )

```

```{r}
returnGini <- function(x){

  f <- paste0(x, list.files(x, pattern = '1k-'))
  mods <- lapply(f, readRDS)
  mods <- parallel::mclapply(mods, randomForest::importance)

  names(mods) <- gsub('1k-.*$', '', basename(f))
  mods <- purrr::map(mods, data.frame) |>
    purrr::map(tibble::rownames_to_column, var = 'Variable') |>
    dplyr::bind_rows(.id = 'Taxon')

  return(mods)
}

giniDecrease <- returnGini(x = '../results/rf_models/')

gd_summ <- giniDecrease |>
  group_by(Taxon) |>
  mutate(
    sumGini = sum(MeanDecreaseGini),
    propDecreaseGini = MeanDecreaseGini/sumGini
  ) |>
  ungroup() |>
  group_by(Variable) |>
  summarize(
    n = n(), 
    mean = mean(propDecreaseGini)
  ) |>
  arrange(-n) |>
  mutate(
    Theme = case_when(
      Variable %in% c(
        'bio1', 'bio10', 'bio11', 'bio12', 'bio18', 'bio19', 'bio4', 'bio5',
        'bio6', 'gdd5', 'gdgfgd5', 'hurs', 'ngd5', 'scd', 'swe', 'vpd') ~ 'Climate',
      Variable %in% c('cfvo_30_60', 'depth2bedrock', 'phh2o_30_60', 'salinity', 
                      'sand_5_15', 'silt_5_15', 'soc_0_5') ~ 'Soil',
      Variable %in% c('aspect', 'cti', 'dem', 'slope', 'tpi', 'tri') ~ 'Topography',
      Variable %in% c('herb', 'lastwild', 'shrub', 'tree') ~ 'Land Cover/Landuse'
      )
    )

ggplot(data = gd_summ, aes(x = n, y = mean, color = Theme, label = Variable)) + 
  geom_smooth(method = 'glm', formula = 'y ~ x', aes(x = n, y = mean), inherit.aes = F, color = 'black') +
  scale_color_manual(
    breaks = c("Climate", "Soil", "Land Cover/Landuse", "Topography"),
    values=c("#773344", "#20A4F3", "#63A375", "#EA7317")
  ) + 
  geom_jitter() + 
  ggrepel::geom_label_repel(max.overlaps = 17, alpha = 0.9) + 
  labs(
    title = 'Relationship between the number of models featuring\n a variable and the variables mean contribution to all models', 
    x = 'Number of models including variable', 
    y = ' Proportional mean decrease in Gini', 
    caption = expression(~italic('Variables become more important as they increase the mean decrease in Gini'))
    ) + 
  theme_classic() + 
  theme(
    axis.line.y = element_blank(), 
    axis.line.x = element_blank(),
    plot.title = element_text(hjust = 0.5), 
    legend.position = 'bottom'
    )

rm(ob, giniDecrease)
```

Of the `r round(sum(recs_removed$totalRecords) * 0.015, 0)` records which were flagged, and manually investigated by an analyst a total of `r sum(recs_removed$recordsRemoved)` were removed. 
Of these `r nrow(recs_removed)` manually reviewed taxa, `r nrow(recs_removed[recs_removed$recordsRemoved==0,])` had no records removed, the mean proportion of records which were removed (based on the entire number of records present) was `r round(mean(recs_removed$propRemoved), 3)`, median = `r round(median(recs_removed$propRemoved), 3)`, with a maximum of `r round(max(recs_removed$propRemoved), 3)`. 
These values are skewed upwards due to a few species which serve as common 'dumping' grounds for difficult to identify taxa in species rich clades.

While the training accuracy scores for the linear-discriminant analysis were high all across the board (mean = `r round(mean(df$ACC_train), 4)`, min = `r round(min(df$ACC_train), 3)`), the realized range on testing values were lower, and reflected values more typical of complex applications (mean = `r round(mean(df$ACC_test), 3)`, median = `r round(median(df$ACC_test), 3)`, min = `r round(min(df$ACC_test), 3)`). 
Of the randomly generated psuedo-absences which underwent linear discriminant analysis for `r nrow(df)` taxa, in only `r nrow(df[df$removedFlagged==0,])` instances did the number of records flagged for removal not exceed the number of records flagged. 
On average `r round(mean(df$propFlagged), 3)` records were flagged per species (median = `r round(median (df$propFlagged), 3)`).

Which variables were the most important rank positions across all models?

How were model evaluation statistics?

@@

```{r}
rm(recs_removed)
```
